---
title: "Accounting for technical batch effects: Response to infection with _Mycobacterium tuberculosis_"
author: "John Blischak"
date: "2018-08-09"
output: workflowr::wflow_html
---

An example of diagnosing and correcting batch effects from one of my own studies
on the response to infection with _Mycobacterium tuberculosis_
([paper](https://www.nature.com/articles/srep16882),
[code](https://bitbucket.org/jdblischak/tb/src),
[data](https://bitbucket.org/jdblischak/tb-data/src)).

## Setup

```{r packages, message=FALSE}
library(dplyr)
library(limma)
library(edgeR)
# Have to load Biobase after dplyr so that exprs function works
library(Biobase)
```

Download data.

```{r data, cache=FALSE}
file_url <- "https://bitbucket.org/jdblischak/tb-data/raw/bc0f64292eb8c42a372b3a2d50e3d871c70c202e/counts_per_sample.txt"
full <- read.delim(file_url, stringsAsFactors = FALSE)
```

Convert to ExpressionSet.

```{r eset}
dim(full)
full <- full[order(full$dir), ]
rownames(full) <- paste(full$ind, full$bact, full$time, sep = ".")

x <- t(full[, grep("ENSG", colnames(full))])
p <- full %>% select(ind, bact, time, extr, rin)
stopifnot(colnames(x) == rownames(p))

eset <- ExpressionSet(assayData = x,
                      phenoData = AnnotatedDataFrame(p))
```

Filter lowly expressed genes.

```{r filter}
keep <- rowSums(cpm(exprs(eset)) > 1) > 6
sum(keep)
eset <- eset[keep, ]
dim(eset)
```

Normalize with TMM.

```{r normalize}
norm_factors <- calcNormFactors(exprs(eset))
exprs(eset) <- cpm(exprs(eset), lib.size = colSums(exprs(eset)) * norm_factors,
                   log = TRUE)
plotDensities(eset, legend = FALSE)
```

Clean up phenotype data frame to focus on early versus late timepoint for this
example.

```{r phenotype}
pData(eset)[, "infection"] <- ifelse(pData(eset)[, "bact"] == "none",
                                     "con", "inf")

pData(eset)[, "time"] <- ifelse(pData(eset)[, "time"] == 4,
                                "early", "late")

pData(eset)[, "batch"] <- sprintf("b%02d", pData(eset)[, "extr"])

table(pData(eset)[, c("time", "batch")])
```

## Remove batch effect

Visualize principal components 1 and 2 for the original data.

```{r pre}
plotMDS(eset, labels = pData(eset)[, "time"], gene.selection = "common")
```

Remove the effect of the technical variables: batch (discrete) and RIN
(continuous; a measure of RNA quality).

```{r removeBatchEffect}
exprs(eset) <- removeBatchEffect(eset, batch = pData(eset)[, "batch"],
                                 covariates = pData(eset)[, "rin"])
```

Visualize principal components 1 and 2 for the corrected data.

```{r post}
plotMDS(eset, labels = pData(eset)[, "time"], gene.selection = "common")
```
