---
title: "Differential expression analysis for studies with 2 groups"
author: "John Blischak"
date: 2018-01-26
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

Analyze proteomics data set that compares 2 groups.

```{r setup, message=FALSE}
library("Biobase")
library("ggplot2")
library("limma")
eset <- readRDS("../data/ch02.rds")
```

## Studies with 2 groups (Video)

Describe the scientific question, the experimental design, and the data
collected for the 2-group study. Introduce the ExpressionSet object that
contains the data. Review the quality control procedures covered in past
Bioconductor courses (specifically comparing distributions and PCA).

## Plot a differentially expressed gene

Create boxplots of a few pre-selected genes (one clearly DE, one
clearly not, and one in between). Use `pData` to select the phenotype
variables and `exprs` to access the expression data.

```{r}
boxplot(exprs(eset)[3, ] ~ pData(eset)$spikein, main = fData(eset)$protein[3])
boxplot(exprs(eset)[5, ] ~ pData(eset)$spikein, main = fData(eset)$protein[5])
boxplot(exprs(eset)[463, ] ~ pData(eset)$spikein, main = fData(eset)$protein[463])
```

## Create a density plot to compare distribution across samples

Use `limma::plotDensities` to confirm that the distribution of gene
expression levels is consistent across the samples.

```{r}
plotDensities(exprs(eset), legend = FALSE)
```

## Create a PCA plot to assess source of variation

Use `prcomp` to compute principal components and then plot PC1 vs. PC2
to confirm that the biological effect is the main source of variation.

```{r}
pca <- prcomp(t(exprs(eset)), scale. = TRUE)
plot(pca)
d <- data.frame(pData(eset), pca$x)
ggplot(d, aes(x = PC1, y = PC2, color = spikein)) +
  geom_point()
```

## limma for differential expression  (Video)

Describe the standard limma workflow. Describe the 2 main techniques
for constructing the linear model: treatment-contrasts versus
group-means parametrizations.

## Create design matrix for treatment-contrasts parametrization

Use `model.matrix` to create a linear model with an intercept and one
binary variable. Use `colSums` to reason about how this relates to the
samples (e.g. the intercept represents the mean across samples because
it is 1 for every sample).

```{r}
design <- model.matrix(~spikein, data = pData(eset))
colSums(design)
```

## Fit and test model for treatment-contrasts parametrization

Use `limma::lmFit`, `limma::eBayes`, and `limma::decideTests` to fit
and test the model.

```{r}
fit <- lmFit(eset, design)
head(fit$coefficients)
fit <- eBayes(fit)
results <- decideTests(fit[, 2])
summary(results)
```

## Create design matrix and contrasts matrix for group-means parametrization

Use `model.matrix` to create a linear model with two binary variables
(and no intercept). Use `colSums` to reason about how this relates to
the samples (e.g. each of the terms represents the mean for its group
of samples because it is the only term that is 1 for those samples).
Use `limma::makeContrasts` to create a contrasts matrix based on this
new linear model.

```{r}
design <- model.matrix(~0 + spikein, data = pData(eset))
colSums(design)
cont_mat <- makeContrasts(spike_effect = spikein25fmol - spikein10fmol,
                          levels = design)
cont_mat
```

## Fit and test model for group-means parametrization

Use `limma::lmFit`, `limma::contrasts.fit`, `limma::eBayes`, and
`limma::decideTests` to fit and test the model. Confirm that the
results are identical to the more traditional linear modelling
approach used previously.

```{r}
fit <- lmFit(eset, design)
head(fit$coefficients)
fit2 <- contrasts.fit(fit, contrasts = cont_mat)
head(fit2$coefficients)
fit2 <- eBayes(fit2)
results <- decideTests(fit2)
summary(results)
```

## Visualizing the results  (Video)

Describe how to access the results with `topTable` and describe the
columns. Demonstrate some common visualizations.

## Create a histogram of p-values

Use `geom_histogram` to plot `P.Value` column from `limma::topTable`.
Ask question to confirm they understand that the p-value distribution
corresponds to the number of differentially expressed genes
identified.

```{r}
topTable(fit2)
stats <- topTable(fit2, number = nrow(fit2), sort.by = "none")
ggplot(stats, aes(x = P.Value)) +
  geom_histogram()
```

## Create a volcano plot

Use `geom_point()` to plot `-log10(P.Value)` vs. `logFC`. Mention
`limma::volcanoPlot` after exercise is completed.

```{r}
stats$significant <- stats$adj.P.Val < 0.05
ggplot(stats, aes(x = logFC, y = -log10(P.Value), color = significant)) +
  geom_point()
```

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
