---
title: "Differential expression analysis for studies with more than 2 groups"
author: "John Blischak"
output: workflowr::wflow_html
---




Analyze RNA-seq data from Hoban et al., 2016, which measured gene expression in
the prefrontal cortex of 3 groups of mice: conventional (con), germ-free (gf),
and germ-free colonized with normal microbiota (exgf).

```{r setup, message=FALSE}
library("Biobase")
library("edgeR")
library("ggplot2")
library("limma")
library("stringr")
dge <- readRDS("../data/ch03.rds")
```

## Studies with more than 2 groups (Video)

Describe the scientific question, the experimental design, and the data
collected for the 3-group study. Review RNA-seq technique and the need to
standardize by library size.

## Calculate the library size

Use `colSums` to calculate the library sizes and run `summary` to appreciate the
variability.

```{r}
lib_size <- colSums(dge$counts)
summary(lib_size)
```

## Calculate counts per million

Use `edgeR::calcNormFactors` and `edgeR::cpm` to calculate TMM-normalized counts
per million.

```{r}
dge <- calcNormFactors(dge)
dge_cpm <- cpm(dge, log = TRUE)
```

## Create density plots of counts versus counts per million

Use `limma::plotDensities` to visualize the distribution of counts and the
distribution of counts per million.

```{r}
plotDensities(dge_cpm, legend = FALSE)
```

## Filtering genes and samples (Video)

Describe the process of filtering lowly expressed genes (show density plots from
previous exercise to demonstrate bimodal distribution) and identifying outlier
samples with PCA. Discuss the nuances of removing outlier samples.

## Remove lowly expressed genes

Remove any gene that does not have at least a log2 cpm > 0 in at least 4 samples
(n = 4 per group). Use `limma::plotDensities` to confirm the distribution of
counts per million is unimodal post-filtering.

```{r}
dge_filt <- dge[rowSums(dge_cpm > 0) >= 4, ]
dge_filt <- calcNormFactors(dge_filt)
dge_filt_cpm <- cpm(dge_filt, log = TRUE)
plotDensities(dge_filt_cpm, legend = FALSE)
```

## Use PCA to identify outlier sample

Use `prcomp` to calculate PCA and then plot PC1 vs. PC2. Remove the exgf sample
that clusters with the gf samples.

```{r}
pca <- prcomp(t(dge_filt_cpm), scale. = TRUE)
plot(pca)
d <- data.frame(dge_filt$samples, pca$x)
ggplot(d, aes(x = PC1, y = PC2, color = group)) +
  geom_text(aes(label = id))

plotMDS(dge_filt)

dge_final <- dge_filt[, dge_filt$samples$id != "exgf1"]
```

## Differential expression for RNA-seq with voom (Video)

Review how to use `model.matrix` and `limma::makeContrasts`. Describe how
`limma::voom` corrects for the mean-variance relationship of the count data.

## Create design matrix for study with 3 groups

Use `model.matrix` to create a linear model with three binary variables
(group-means parametrization).

```{r}
design <- model.matrix(~0 + group, data = dge_final$samples)
colnames(design) <- str_replace(colnames(design), "group", "")
colSums(design)
```

## Create contrasts matrix for study with 3 groups

Use `limma::makeContrasts` to test all pairwise comparisions.

```{r}
cont_mat <- makeContrasts(gf.con = gf - con,
                          exgf.con = exgf - con,
                          exgf.gf = exgf - gf,
                          levels = design)
cont_mat
```

## Account for mean-variance relationship with voom

Use `limma::voom` to calculate weights for the linear model to account for the
mean-variance relationship of the counts. Set `plot = TRUE` to visualize the
mean-variance trend. Explore the EList object returned to confirm that the
expression values in `E` are simply the counts per million.

```{r}
dge_final <- calcNormFactors(dge_final)
v <- voom(dge_final, design, plot = TRUE)
plotDensities(v$E, legend = FALSE)
dge_final_cpm <- cpm(dge_final, log = TRUE)
plot(v$E[, 1], dge_final_cpm[, 1])
```

## Fit and test model for study with 3 groups

Use `limma::lmFit`, `limma::contrasts.fit`, `limma::eBayes`, and
`limma::decideTests` to fit and test the model. Use `limma::vennDiagram` to
visualize overlap of differentially expressed genes.

```{r}
fit <- lmFit(v, design)
fit2 <- contrasts.fit(fit, contrasts = cont_mat)
fit2 <- eBayes(fit2)
results <- decideTests(fit2)
vennDiagram(results)
```


